# Datasets and plots {-#datasets}

```{r, echo = FALSE}
library(mdatools)
```

The package uses standard representation of the data in R: data frames, matrices and vectors. However, there are several additional methods and attributes, which makes the use of the datasets a bit more more efficient. There is also a support for images. But if you are used to simple datasets and standard procedures and do not want any complications, you can simply skip this chapter.

The package also uses its own set of plotting tools, which is a sort of an add-on for the R basic plotting system, extending its possibilities. From this point of view, learning how these tools work will simplify understanding of model plots a lot. The main improvements comparing to the basic plotting system are:

1. Much easier way to make plots with groups of objects (points, lines, bars, etc.)
2. Much easier way of adding legend to the group plots.
3. Much easier way of adding labels to data points, bars, etc.
4. Automatic axis limits when a plot contains several groups of objects.
5. Possibility to color points and lines according to values of a specific numerical variable of a factor.
6. Several built in color pallettes and an easy way to use user specific set of colors.
7. Much more!

This chapter explains most of the details.

## Attributes and factors {-#datasets--attributes-and-factors}

This section tells how to extend the functionality of the package by using attributes assigned to datasets and how methods implemented in the package deal with factors.

### Attributes for plots  {-}

The plot attributes will be explained very briefly here, you can find much more details in the next two sections. The idea is to provide some names and values to the data, which can be used later e.g. for making labels and titles on the plots. When dataset is used to create a model (e.g. PCA) all results representing objects (e.g. scores, distances, etc.) will inherit the row specific attributes and all results related to variables (e.g. loadings) will inherit column specific attributes.

The main attributes for plots are following:

----------------------------------------------------------------------------
Attribute         Meaning
----------        ----------------------------------------------------------
 `name`           name of a dataset (used for plot parameter `main`).

 `xaxis.name`     if variables/data columns represent the same property, e.g. wavelength,
                  the property name can be defined by this attribute (used for plot axes labels).

 `yaxis.name`     if objects/data rows represent the same property, e.g. reaction time,
                  the property name can be defined by this attribute (used for plot axes labels).

 `xaxis.values`   a vector of values, which correspond to the columns (e.g. for
                  spectroscopic data it can be a vector with wavelength or wavenumbers).

 `yaxis.values`   a vector of values, which correspond to the rows (e.g. for
                  kinetic data it can be a vector with time or temperature values).

----------------------------------------------------------------------------


Here is a very simple example.

```{r, fig.width = 9, fig.height = 5}

# generate data as matrix with two rows and 63 columns
t = -31:31
d = rbind(sin(t/10), cos(t/10)) + 1

# add name of the dataset
attr(d, "name") = "Concentration profiles"

# add name and values for a property representing the columns (time of reaction)
# we make values for x-axis to span from 0 to 620 seconds
attr(d, "xaxis.name") = "Time, s"
attr(d, "xaxis.values") = (t + 31) * 10

# make a line plot for the data
mdaplot(d, type = "l")
```

As you can notice, both the label and the ticks for x-axis correspond to the attributes we defined. As well as the main title of the plot. As we mentioned above these attributes will be inherited and e.g. PCA loadings will have the same x-axis if you decide to make a line plot for the loadings.

See more details in the [section][Simple plots] about plots.

### Special methods for data transformations {-}

Since data objects in R loose all user specified attributes when e.g. we transpose them or taking a subset, it was decided to write several methods, which would process attributes correctly. They also adjust indices of excluded rows and columns when user takes a subset or merge two data objects together. When data matrix is transposed the corresponding method will swap the x- and y- attributes. All methods with a brief description are listed in the table below.

-------------------------------------------------------------------------------------------------
Method                                Description
----------------                      -----------------------------------------------------------
`mda.show(data)`                      Show data object without excluded elements

`mda.t(data)`                         Transpose data object

`mda.cbind(data1, data2, ...)`        Merge several datasets by columns

`mda.rbind(data1, data2, ...)`        Merge several datasets by rows

`mda.subset(data1, subset, select)`   Take a subset of data object (`subset` is numeric indices,
                                      names or logical values for rows, `select` — the same
                                      for columns)

`attrs = mda.getattr(data)`           Return all user specific attributes from an object

`data = mda.setattr(data, attrs)`     Assign user specific attributes to an object

------------------------------------------------------------------------------------------------

To avoid any problems with arguments use these functions instead of the traditional ones, e.g. `mda.subset()` instead of `subset()`, when available.


### Data frames with factors {-}

All methods, implemented in the package, work with matrices, therefore, if a user provides data values as data frame, it is converted to matrix. It is also possible to provide data frames with one or several factor columns. In this case all factors will be converted to dummy variables with values 0 and 1. You can also do it manually, by using function `prep.df2mat()` as this is shown in an example below.

Let us first create a simple data with a factor column.

```{r}
h = c(180, 175, 165, 190, 188)
c = as.factor(c("Gray", "Green", "Gray", "Green", "Blue"))
d = data.frame(Height = h, Eye.color = c)
show(d)
```

And this is the result of converting it to a matrix.

```{r}
d.mat = mda.df2mat(d)
show(d.mat)
```

The number of dummy variables by default is the number of levels minus one. You can change this by using argument `full = TRUE` is it is shown in the example below.

```{r}
d.mat = mda.df2mat(d, full = TRUE)
show(d.mat)
```

It is important to have level labels in all factor columns of the same data frame unique, as they are used for names of the dummy variables (e.g. you should not have two factors with the same level name). If a factor is hidden it will be just converted to numeric values and remain excluded from modelling.

## Excluding rows and columns  {-#datasets--excluding-rows-and-columns}

Sometimes it is necessary to hide or to exclude a particular row or a particular column from a data, so they will not be shown on plots and will not be taken into account in modelling, but without removing them physically. So the excluded rows and columns are still there but are not treated in a usual way by the *mdatools* methods. In *mdatools* it is possible by using the following functions:

------------------------------------------------------------------------------------------------------
Function                   Description
------------               ---------------------------------------------------------------------------
`mda.exclrows(x, ind)`     Exclude (hide) all rows specified by variable `ind`, which can be a vector
                           with rows indices, names or a logical vector.

`mda.exclcols(x, ind)`     Exclude (hide) all columns specified by variable `ind`, which can be a vector
                           with columns indices, names or a logical vector.

------------------------------------------------------------------------------------------------------

The mechanism is very simple, the indices of the rows or columns, which must be excluded, are saved into special attributes, which then is recognized by all methods implemented in `mdatools`. Standard R functions will ignore the attributes.

Here is a simple example. Let's create a dataset first (it can be either matrix or a data frame)

```{r}
Height = c(180, 175, 165, 190, 188)
Weight = c(78, 79, 60, 99, 80)
Shoesize = c(44, 39, 35, 45, 44)
d = cbind(Height, Weight, Shoesize)
rownames(d) = paste0("O", 1:5)
show(d)
```

Now let's exclude rows 3 and 4 and then column with name "Weight" from the data.

```{r}
d = mda.exclrows(d, 3:4)
d = mda.exclcols(d, "Weight")
show(d)
```

As you can see, all the values are still there, but there are two new attributes, `exclcols` and `exclrows`. To avoid any issues do not change the values of the attributes manually, always use the functions above.

Now if you make a plot using function `mdaplot()` you will see only three points by default, because the other two a hidden. Also plot will be made for columns 1 and 3, because the second column (Weight) was also hidden. See the code and the result below:

```{r}
mdaplot(d, show.labels = TRUE)
```

You can include the hidden columns and rows back by using `mda.inclcols()` and `mda.inclrows()`.

## Simple plots {-#datasets--simple-plots}

As it was already mentioned, *mdatools* has its own functions for plotting with several extra options not available in basic plot tools. These functions are used to make all plots in the models and results (e.g. scores, loadings, predictions, etc.) therefore it can be useful to spend some time and learn the new features (e.g. coloring data points with a vector of values or using manual ticks for axes). But if you are going to make all plots manually (e.g. using *ggplot2*) you can skip this and the next sections.

In this section we will look at how to make simple plots from your data objects. Simple plots are scatter (`type = "p"`), density-scatter (`type = "d"`), line (`type = "l"`), line-scatter (`type = "b"`), bar (`type = "h"`) or errorbar (`type = "e"`) plots made for a one set of objects. All plots can be created using the same method `mdaplot()` by providing a whole dataset as a main argument. Depending on a plot type, the method "treats" the data values differently.

This table below contains a list of parameters for `mdaplot()`, which are not available for traditional R plots. In this section we will describe most of the details using simple examples.

--------------------------------------------------------------------------------
Parameter            Description
------------         -----------------------------------------------------------
`cgroup`             a vector of values (same as number of rows in data) used to colorize plot
                     objects with a color gradient.

`colmap`             color map for the color gradient (possible values are `'default'`, `'gray'`
                     or a vector with colors).

`show.colorbar`      when color grouping is used, `mdaplot()` shows a color bar legend, this
                     parameter allows to turn it off.

`show.labels`        logical parameter showing labels beside plot objects (points, lines, etc).
                     Size and color of labels can be adjusted using parameters `lab.cex` and `lab.col`.

`labels`             parameter telling what to use as labels (by default row names, but can also
                     be indices or manual values).

`lab.col`            color for the labels.

`lab.cex`            font size for the labels (as a scale factor).

`xticks`             vector with numeric values to show the x-axis ticks at.

`yticks`             vector with numeric values to show the y-axis ticks at.

`xticklabels`        vector with labels (numbers or text) for the x-ticks.

`yticklabels`        vector with labels (numbers or text) for the y-ticks.

`xlas`               an integer between 0 and 3 telling at which angle the x-tick labels
                     have to be shown.

`ylas`               an integer between 0 and 3 telling at which angle the y-tick labels have
                     to be shown.

`show.axes`          logical, if `TRUE`, function will make a new plot, if `FALSE`, add the plot
                     objects to a previous one.

`show.lines`         a vector with two numbers — position of horizontal and vertical lines on a
                     plot (e.g. coordinate axes).

`show.grid`          logical, show or not a grid. It places grid behind the plot object in contrast
                     to conventional `grid()` method. Use `grid.lwd` and `grid.col` parameters to
                     adjust the grid look.

`show.excluded`      logical, show or not points or lines corresponded to the excluded rows.

`opacity`            opacity of colors in range 0...1 (applied to all colors of current plot).

--------------------------------------------------------------------------------

### Scatter plots {-}

We will use `people` dataset for illustration how scatter plots work (see `?people` for details).

```{r}
data(people)
attr(people, "name") = "People"
attr(people, "xaxis.name") = "Variables"
attr(people, "yaxis.name") = "Persons"
```

For scatter plots the method takes first two columns of a dataset as x and y vectors. If only one column is available `mdaplot()` uses it for y-values and generate x-values as an index for each value.

```{r, fig.width = 9, fig.height = 5}
par(mfrow = c(1, 2))

# show plot for the whole dataset (columns 1 and 2 will be taken)
mdaplot(people, type = "p")

# subset the dataset and keep only columns 6 and 7 and then make a plot
mdaplot(mda.subset(people, select = c(6, 7)), type = "p")
```

All parameters, available for the standard `points()` method will work with `mdaplot()` as well. Besides that, you can colorize points according to some values using a color gradient. By default, the gradient is generated using one of the diverging color schemes from [colorbrewer2.org](http://colorbrewer2.org), but this can be changed using parameter `colmap` as it is shown below.

```{r, fig.width = 9, fig.height = 9}
par(mfrow = c(2, 2))

# show Height vs Weight and color points by the Beer consumption
mdaplot(people, type = "p", cgroup = people[, "Beer"])

# do the same but do not show colorbar
mdaplot(people, type = "p", cgroup = people[, "Beer"], show.colorbar = FALSE)

# do the same but use grayscale color map
mdaplot(people, type = "p", cgroup = people[, "Beer"], colmap = "gray")

# do the same but using colormap with gradients between red, yellow and green colors
mdaplot(people, type = "p", cgroup = people[, "Beer"], colmap = c("red", "yellow", "green"))
```

If the vector with values for color grouping is a factor, level labels will be shown on a colorbar legend and there will be a small margin between bars.

```{r, fig.width = 9, fig.height = 5}
# make a factor using values of variable Sex and define labels for the factor levels
g = factor(people[, "Sex"], labels = c("Male", "Female"))

par(mfrow = c(1, 2))
mdaplot(people, type = "p", cgroup = g)
mdaplot(people, type = "p", cgroup = g, colmap = "gray")
```

If you use point characters from 21 to 25 (the ones which allow to specify both color of border and
background of the marker symbol), then the `cgroup` is applied to color of the borders of the symbols by default.
If you want to apply it to background color, use logical parameter `pch.colinv` for that as shown below:

```{r, fig.width = 9, fig.height = 5}
par(mfrow = c(1, 2))

# default way - color grouping is used for borders and "bg" for background
mdaplot(people, type = "p", cgroup = people[, "Beer"], pch = 21, bg = "white")

# inverse - color grouping is used for background and "bg" for border
mdaplot(people, type = "p", cgroup = people[, "Beer"], pch = 21, bg = "white", pch.colinv = TRUE)
```

Another useful option is adding labels to the data points. By default row names will be taken for the labels but you can specify a parameter `labels`, which can be either a text (`"names"` or `"indices"`) or a vector with values to show as labels. Color and size of the labels can be adjusted.

```{r, fig.width = 9, fig.height = 9}
par(mfrow = c(2, 2))

# by default row names will be used as labels
mdaplot(people, type = "p", show.labels = TRUE)

# here we tell to use indices as labels instead
mdaplot(people, type = "p", show.labels = TRUE, labels = "indices")

# here we use names again but change color and size of the labels
mdaplot(people, type = "p", show.labels = TRUE, labels = "names", lab.col = "red", lab.cex = 0.5)

# finally we provide a vector with manual values to be used as the labels
mdaplot(people, type = "p", show.labels = TRUE, labels = paste0("O", seq_len(nrow(people))))
```

You can also manually specify axis ticks and tick labels. The labels can be rotated using parameters `xlas` and `ylas`, see the examples below. It is important though, that if you provide manual values for the tick labels you must also provide a vector of values with the positions the labels should be shown at. And, of course, the two vectors must
have the same number of values.

```{r, fig.width = 9, fig.height = 9}
par(mfrow = c(2, 2))

# manual values and tick labels for the x-axis
mdaplot(people, xticks = c(165, 175, 185), xticklabels = c("Small", "Medium", "Hight"))

# same but with rotation of the tick labels
mdaplot(people, xticks = c(165, 175, 185), xticklabels = c("Small", "Medium", "Hight"), xlas = 2)

# manual values and tick labels for the y-axis
mdaplot(people, yticks = c(55, 70, 85), yticklabels = c("Light", "Medium", "Heavy"))

# same but with rotation of the tick labels
mdaplot(people, yticks = c(55, 70, 85), yticklabels = c("Light", "Medium", "Heavy"), ylas = 2)
```

If both axis labels and rotated axis ticks have to be shown, you can adjust plot margins and position of the label using `par()` function and `mtext()` for positioning axis label manually.

```{r, fig.width = 9, fig.height = 5}
par(mfrow = c(1, 2))

# change margin for bottom part
par(mar = c(6, 4, 4, 2) + 0.1)
mdaplot(people, xticks = c(165, 175, 185), xticklabels = c("Small", "Medium", "Hight"),
        xlas = 2, xlab = "")
mtext("Height", side = 1, line = 5)

# change margin for left part
par(mar = c(5, 6, 4, 1) + 0.1)
mdaplot(people, yticks = c(55, 70, 85), yticklabels = c("Light", "Medium", "Heavy"),
        ylas = 2, ylab = "")
mtext("Weight", side = 2, line = 5)
```

```{r, echo = FALSE}
par(mar = c(5, 4, 4, 2) + 0.1)
```

There is also a couple of other parameters, allowing to show/hide grid as well as show horizontal and vertical lines on the plot (axes limits will be adjusted correspondingly).

```{r, fig.width = 9, fig.height = 5}
par(mfrow = c(1, 2))
mdaplot(people, show.grid = FALSE, show.lines = c(170, 65))
mdaplot(people, show.lines = c(220, NA))
```

Function `mdaplot()` can also return plot series data, which can be used for
extra options. For example, in case of scatter plot you can add confidence ellipse or convex
hull for data points. To do this, points must be color grouped by a factor as shown below. For confidence ellipse you can specify the confidence level (default 0.95).

```{r, fig.width = 9, fig.height = 5}

# define a factor using values of variable Sex and simple labels
g = factor(people[, "Sex"], labels = c("M", "F"))

par(mfrow = c(1, 2))

# make a scatter plot grouping points by the factor and then show convex hull for each group
p = mdaplot(people, cgroup = g)
plotConvexHull(p)

# make a scatter plot grouping points by the factor and then show 90% confidence intervals
p = mdaplot(people, cgroup = g)
plotConfidenceEllipse(p, conf.level = 0.90)
```

In case when number of data points is large (e.g. when dealing with images, where every pixel is a data point), using density plot is a good alternative to conventional scatter plots. The plot does not show all data points but instead split the whole plot space into small hexagonal regions and use color gradient for illustration a density of the points in each region. This approach is known as *hexagonal binning*. To create a density plot simply use `type="d"`. You can also specify color map and number of bins along each axes (`nbins`).

The code below show an example of using density plots for 100000 data points with x and y values taken from normally distributed population.

```{r, fig.width = 9, fig.height = 9}
x = rnorm(100000)
y = rnorm(100000)
d = cbind(x, y)

par(mfrow = c(2, 2))
mdaplot(d, type = "d")
mdaplot(d, type = "d", nbins = 80)
mdaplot(d, type = "d", colmap = "gray")
mdaplot(d, type = "d", nbins = 80, colmap = c("blue", "red"))
```

### Line plots {-}

When line plot is created, the `mdatools()` shows a line plot for every row of the provided dataset. So if data set has more than one row, the plot will show a bunch of lines having same properties (color, type, etc). This is particularly useful when working with signals and spectroscopic data. In this subsection we will use simulated UV/Vis spectra from `simdata`. See `?simdata` for more details about this set.

```{r}
data(simdata)

# get the calibration spectra, wavelength and
# concentration of first component as separate variables
spectra = simdata$spectra.c
wavelength = simdata$wavelength
conc = simdata$conc.c[, 1]

# add names as attributes
attr(spectra, "name") = "UV/Vis spectra"
attr(spectra, "xaxis.name") = "Band index"
```

Here are simple examples of how to make the line plots.

```{r, fig.width = 9, fig.height = 9}
par(mfrow = c(2, 1))
mdaplot(spectra, type = "l")
mdaplot(spectra, type = "l", col = "darkgray", lty = 2)
```

Most of the parameters described for scatter plots will work for the line plots as well. For example, you can colorise the lines by using a vector with some values (in the example below I use concentration of one of the chemical components).

```{r, fig.width = 9, fig.height = 5}
par(mfrow = c(1, 1))
mdaplot(spectra, type = "l", cgroup = conc)
```

And of course you can use the attributes, allowing to provide manual x-values — `'xaxis.values'` (similar parameter for y-values is `'yaxis.values'`). In the example below we show the spectra using wavelength in nm and wavenumbers in inverse cm.

```{r, fig.width = 9, fig.height = 9}
par(mfrow = c(2, 1))

attr(spectra, "xaxis.name") = expression("Wavenumbers, cm"^-1)
attr(spectra, "xaxis.values") = 10^7/wavelength
mdaplot(spectra, type = "l")

attr(spectra, "xaxis.name") = "Wavelength, nm"
attr(spectra, "xaxis.values") = wavelength
mdaplot(spectra, type = "l")
```

When you provide such data to any model methods (e.g. PCA, PLS, etc), then all variable related results (loadings, regression coefficients, etc.) will inherit this attribute and use it for making line plots.

### Bar and errorbar plots {-}

Bar plot is perhaps the simplest as it shows values for the first row of the data as bars. Let us start with simple dataset, where we have a matrix with explained variance of a data from e.g. PCA decomposition.

```{r, fig.width = 9, fig.height = 5}
# make a simple two rows matrix with values
d = rbind(
   c(20, 50, 60, 90),
   c(14, 45, 59, 88)
)

# add some names and attributes
colnames(d) = paste0("PC", 1:4)
rownames(d) = c("Cal", "CV")
attr(d, "xaxis.name") = "Components"
attr(d, "name") = "Explained variance"

par(mfrow = c(1, 2))

# make a default bar plot
mdaplot(d, type = "h")

# make a bar plot with manual xtick labels, color and labels for data values
mdaplot(d, type = "h", xticks = seq_len(ncol(d)), xticklabels = colnames(d), col = "red",
   show.labels = TRUE, labels = "values", xlas = 2, xlab = "", ylab = "Variance, %")
```

As you can notice, the values from the second rows were ignored, as bar plot always takes the first row.

Errorbar plot, in contrast, always expect data to have two or three rows. The first row is the origin points of the error bars, second row is the size of the bottom error bar and the third row is the size of the top error bar. If data has only two rows the both parts will be symmetric related to the origin.

```{r, fig.width = 9, fig.height = 5}

# generate some mean and standard deviation values and combine them as rows of a matrix
d = rbind(
   c(20, 60, 70, 75),
   c(2, 5, 4, 3)
)

# add names and attributes
rownames(d) = c("Mean", "Std")
colnames(d) = paste0("PC", 1:4)
attr(d, 'name') = "Statistics"


# show the plots
par(mfrow = c(1, 2))
mdaplot(d, type = "e")
mdaplot(d, type = "e", xticks = seq_len(ncol(d)),
   xticklabels = colnames(d), col = "red", xlas = 2, xlab = "")
```

All simple plots can be combined together on the same axes. In this case, first plot is created as usual and all other plots have to be created with option `show.axes = FALSE` as it is shown below. It must be noted that in this case axes limits have to be set manually when creating the first plot.

```{r, fig.width = 9, fig.height = 5}
par(mfrow = c(1, 2))

mdaplot(mda.subset(d, 1), type = "h", col = "lightgray")
mdaplot(d, type = "e", show.axes = FALSE, pch = NA)

mdaplot(mda.subset(d, 1), type = "b")
mdaplot(d, type = "e", show.axes = FALSE)
```

In the next section we will discuss plots for several groups of objects (rows).

## Plots for groups of objects {-#datasets--group-plots}

The package has another method for creating plots, `mdaplotg()`, which aims at making plots for groups of objects. It can be several groups of points, lines or bars, where every group has its own attributes, such as color, marker, line type and width, etc. There is a simple criterion to distinguish between the simple and group plots: group plots usually need a legend and simple plots — not. The `mdaplotg()` method allows to do a lot of things (e.g. split data into groups, add a legend and labels, etc) much easier and this section will show most of the details.

There are three ways to provide data sets for making the group plots. Let's discuss them first and then talk about some extra features.

### One matrix or data frame {-}

If dataset is a matrix or a data frame, `mdaplotg()` will make a line, scatter-line or a bar plot, considering every row as a separate group. This can be useful, when, for example, you want to show how explained variance depends on a number of components for calibration and test set, or how loadings for the first two components look like.

If you want to change any parameters, like `pch`, `lty`, `lwd`, `col` or similar you need to provide either a vector with values for each group or one value for all groups. Axis limits, ticks, ticklabels, etc. can be defined similarly to the simple plots. Here are some examples.

```{r, fig.width = 9, fig.height = 9}
# let's create a simple dataset with 3 rows
p = rbind(
   c(0.40, 0.69, 0.88, 0.95),
   c(0.34, 0.64, 0.81, 0.92),
   c(0.30, 0.61, 0.80, 0.88)
)

# add some names and attributes
rownames(p) = c("Cal", "CV", "Test")
colnames(p) = paste0("PC", 1:4)
attr(p, "name") = "Cumulative variance"
attr(p, "xaxis.name") = "Components"

# and make group plots of different types
par(mfrow = c(2, 2))
mdaplotg(p, type = "l")
mdaplotg(p, type = "b")
mdaplotg(p, type = "h", xticks = 1:4)
mdaplotg(p, type = "b", lty = c(1, 2, 1), col = c("red", "green", "blue"), pch = 1,
         xticks = 1:4, xticklabels = colnames(p))
```

As you can see, `mdaplotg()` automatically created the legend and set colors, line parameters, etc. correctly. You can change position of the legend using same names as for basic `legend()` command from R, or hide it using parameter `show.legend = FALSE`, as it is shown below.

```{r, fig.width = 9, fig.height = 9}
par(mfrow = c(2, 2))
mdaplotg(p, type = "l", legend.position = "top")
mdaplotg(p, type = "b", legend.position = "bottomleft")
mdaplotg(p, type = "h", legend.position = "bottom")
mdaplotg(p, type = "b", show.legend = FALSE)
```

Group plot also allows to show labels, in this case they can be either values, names or indices of the columns.

```{r, fig.width = 9, fig.height = 9}
par(mfrow = c(2, 2))
mdaplotg(p, type = "l", show.labels = TRUE)
mdaplotg(p, type = "b", show.labels = TRUE, labels = "indices")
mdaplotg(p, type = "h", show.labels = TRUE, labels = "values")
mdaplotg(p, type = "b", show.labels = TRUE, labels = "values")
```

### List with matrices or data frames {-}

In this case every element of the list will be treated as a separate group. This way allows to make scatter plots as well and line plots with several lines in each group. Barplot can be also made but in this case first row from each datasets will be used. If you use names when create the list, the names will be taken as legend labels, otherwise method will look at attribute `"name"` for each data set.

In the example below we split *People* data to males and females and show the group plots.

```{r, fig.width = 9, fig.height = 9}

# load data and exclude column with income
data(people)
people = mda.exclcols(people, "Income")

# use values of sex variable to split data into two subsets
sex = people[, "Sex"]
m = mda.subset(people, subset = sex == -1)
f = mda.subset(people, subset = sex ==  1)

# combine the two subsets into a named list
d = list(male = m, female = f)

# make plots for the list
par(mfrow = c(2, 2))
mdaplotg(d, type = "p")
mdaplotg(d, type = "b")
mdaplotg(d, type = "h")
mdaplotg(d, type = "b", lty = c(1, 2), col = c("red", "blue"), pch = 1)
```

Most of the things described in the previous subsection will work similarly for this case. We will just add a bit more details on how labels and excluded rows are processed for the scatter plots. By default labels are row names or indices. In `mdaplotg()` you can not provide vector with manual values, so the best way to change them is to assign them as the row names. Indices are unique within each group, so if you have, e.g. three groups of points, there will be three points with index "1", three with "2", etc.

### Use factors to split a dataset into groups {-}

One more way to split data set into groups is to provide one or several factor columns using argument `groupby`. In this case `mdaplotg()` will find all combinations of the factor levels and split rows of dataset to the corresponding groups. In the example below we use variables *Region* and *Sex* to make plots for four groups.

It is assumed that you already loaded the Poeple data (from previous examples) and excluded the Income column.

```{r, fig.width = 9, fig.height = 9}
sex = factor(people[, "Sex"], labels = c("M", "F"))
reg = factor(people[, "Region"], labels = c("S", "M"))
groups = data.frame(sex, reg)

par(mfrow = c(2, 2))
mdaplotg(people, type = "p", groupby = groups)
mdaplotg(people, type = "l", groupby = groups)
mdaplotg(people, type = "b", groupby = groups)
mdaplotg(people, type = "h", groupby = groups)
```

All parameters, described before, will work the same way in this case.


## Working with images {-#datasets--images}

The package also supports images, including hyperspectral images, however they have to be transformed into datasets. The idea is very simple, we keep information about image pixels in an unfolded form, as a matrix, and use attributes `width` and `height` to reshape the data when we need to show it as an image.

There are three methods that make this procedure easier: `mda.im2data()`, `mda.data2im()` and `imshow()`. The first converts an image (represented as 3-way array) to a data set, second does the opposite and the third takes dataset and shows it as an image. In the code chunk below you will see several examples how the methods work.

We will use a dataset `image` available in the package. It is a 3-way array of numbers, if you want to work with e.g. JPEG, PNG or other standard image files you can load them using specific packages ([jpeg](https://cran.r-project.org/web/packages/jpeg/index.html),
[png](https://cran.r-project.org/web/packages/png/index.html)).

```{r, echo = FALSE}
library(mdatools)
```

```{r, fig.width = 9, fig.height = 9}
data(pellets)

# convert image to a data matrix and add some attributed
d = mda.im2data(pellets)
colnames(d) = c("Red", "Green", "Blue")
attr(d, "name") = "Image"

# show data values
mda.show(d, 10)

# show separate channels and the whole image in plots
par(mfrow = c(2, 2))
imshow(d, 1)
imshow(d, 2)
imshow(d, 3)
imshow(d, 1:3)
```

By default image for one channels is shown using jet color palette for intensities, but you can also use gray colors, palette from *colorbrewer2* as well as your own.

```{r, fig.width = 9, fig.height = 9}
par(mfrow = c(2, 2))
imshow(d, 1)
imshow(d, 1, colmap = "gray")
imshow(d, 1, colmap = heat.colors(256))
imshow(d, 1, colmap = colorRampPalette(c("red", "green"))(256))
```

You can work with the image values as normal dataset and show scatter, line plots, calculate statistics, etc.

```{r, fig.width = 9, fig.height = 5}
par(mfrow = c(1, 2))
mdaplot(d, type = "p")
mdaplot(d, type = "l")
```

However, it will take some time to show these plots as this image has several hundreds of thousands pixels, a faster alternative can be the use density plot based on hexagonal binning. Use `type = "d"` for this as shown below. Number of bins can be adjusted by using parameter `nbins`.

```{r, fig.width = 9, fig.height = 5}
par(mfrow = c(1, 2))
mdaplot(d, type = "d")
mdaplot(mda.subset(d, select = c("Red", "Blue")), type = "d", nbins = 40)
```

Another useful thing is to set some of the pixels as background. The background pixels are removed from the image dataset physically, there is no way to get them back (in cotrast to excluded rows/pixels). It can be particularly useful when working with e.g. geocorrected hyperspectral images, where, often, many pixels have NA values and there is no need to keep them in memory. To set pixels as background you need to use method `mda.setimbg()` with either pixel indices or vector with logical values as it is shown below.

```{r, fig.width = 9, fig.height = 5}

# original size
show(dim(d))

# set red epixels as background and show new size
d = mda.setimbg(d, d[, "Red"] > 100)
show(dim(d))

# show image with background pixels
par(mfrow = c(1, 2))
imshow(d, 1)
imshow(d, 1:3)
```

All image related attributes are inherited by all object/rows related results, e.g. scores, residuals, predicted values and classes, etc. This means if you provide an image to any modelling method, you can visualise the corresponding results also as an image. Some examples will be shown in
[chapter about PCA][Principal component analysis].

