# SIMCA classification {-}

SIMCA (Soft Independent Modelling of Class Analogy) is a simple but efficient one-class classification method mainly based on PCA. The general idea is to create a PCA model using data for samples/objects belonging to a class and classify new objects based on how good the model can fit them. The decision is made using two residual distances — $Q$ and $T^2$ and corresponding critical limits. Critical limits computed for both distances (or their combination) are used to cut-off the strangers (extreme objects) and accept class members with a pre-define expected ratio of false negatives ($\alpha$). All details about which method for critical limits implemented in this package can be found [here][Residuals and critical limits]. It must be noted that version *0.9.0* brings a lot if improvements and new features related to the critical limits calculation and SIMCA classification, so it can be a good idea to read this chapter first.

The classification performance can be assessed using true/false positives and negatives and statistics, showing the ability of a classification model to recognize class members (*sensitivity* or true positive rate) and how good the model is for identifying strangers (*specificity* or true negative rate). In addition to that, model also calculates a percent of misclassified objects. All statistics are calculated for calibration and validation (if any) results, but one must be aware that specificity can not be computed without objects not belonging to the class and, therefore, calibration and cross-validation results in SIMCA do not have specificity values. 

It must be also noted that any SIMCA model or result is also a PCA object and all plots, methods, statistics, available for PCA, can be used for SIMCA objects as well. 

## Calibration and validation {-}

The model calibration is similar to PCA, but there are several additional arguments, which are important for classification. First of all it is a class name. Class name is a string, which can be used later e.g. for identifying class members for testing. The second important argument is a level of significance, `alpha`. This parameter is used for calculation of statistical limits and can be considered as probability for false negatives. The default value is 0.05. Finally the parameter `lim.type` allows to select the method for compuring critical limits for the residuals, as it is described in the PCA chapter.

In this chapter as well as for describing other classification methods we will use a famous Iris dataset, available in R. The dataset includes 150 measurements of three Iris species: *Setosa*, *Virginica* and *Versicola*. The measurements are length and width of petals and sepals in cm. Use `?iris` for more details. 

Let's get the data and split it to calibration and test sets.

```{r}

data(iris)
head(iris)

# generate indices for calibration set
idx = seq(1, nrow(iris), by = 2)

# split the values
X.c = iris[idx, 1:4]
c.c = iris[idx, 5, drop = F]

X.t = iris[-idx, 1:4]
c.t = iris[-idx, 5, drop = F]
```

Now, because for calibration we need only objects belonging to a class, we will split the `X.c` into three matrices — one for each species. The data is ordered by the species, so it can be done relatively easy by taking every 25 rows.

```{r}
X.set = X.c[1:25, ]
X.ver = X.c[26:50, ]
X.vir = X.c[51:75, ]
```

Let's start with creating a model for class *Versicolor* and exploring available statistics and plots. We will use full cross-validation to validate the results. In this case we do not change method for critical limits, so the default option (`jm`) is used.

```{r}
library(mdatools)
m = simca(X.ver, 'versicolor', ncomp = 3, cv = 1)
summary(m)
```

Let's look at plots and start with summary plot.

```{r, fig.width = 9, fig.height = 8}
plot(m)
```

The plot is very similar to what we seen for PCA model, the only difference is that it shows *modelling power* instead of loadings. Modelling power is a measure of contribution of each variable to the model and varies from 0 to 1. Usually variables with modelling power below 0.1 are considered as irrelevant.

Let's give a closer look at the residuals plot with different values for alpha (we will keep number of components equal to three in all cases).

```{r, fig.width = 9, fig.height = 8}
m1 = simca(X.ver, 'versicolor', ncomp = 3, cv = 1, alpha = 0.01)
m2 = simca(X.ver, 'versicolor', ncomp = 3, cv = 1, alpha = 0.05)
m3 = simca(X.ver, 'versicolor', ncomp = 3, cv = 1, alpha = 0.10)
m4 = simca(X.ver, 'versicolor', ncomp = 3, cv = 1, alpha = 0.15)

par(mfrow = c(2, 2))
plotResiduals(m1)
plotResiduals(m2)
plotResiduals(m3)
plotResiduals(m4)
```

As you can see, using alpha = 0.01 reduced number of false negatives to zero, as the acceptance limits became larger, while alpha = 0.15 gives a lot of incorrectly rejected class members. It must be noted, that decreasing alpha will also lead to a larger number of false positives, which we can not see in this case.

### Predictions and validation with a test set {-}

When model is ready one can test it using a new test set with know classes. In this case we will use objects from all three species and be able to see how good the model performs on strangers (and calculate the specificity). In order to do that we will provide both the matrix with predictors, `X.t`, and a vector with names of the classes for corresponding objects/rows (`c.t`). The values with known classes in this case can be:

* a vector with text values (names)
* a factor using the names as labels
* a vector with logical values (`TRUE` for class members and `FALSE` for strangers)

In our case we have a factor. Instead of creating a new model and providing the values as test set we will make predictions instead.

```{r}
res = predict(m, X.t, c.t)
summary(res)
```

In this case we see a more detailed statistics with true/false positives and negatives, specificity and sensitivity. The performance statistics can be also shown as plots.

```{r, fig.width = 9, fig.height = 8}
par(mfrow = c(2, 2))
plotSpecificity(res)
plotSensitivity(res)
plotMisclassified(res)
plotPerformance(res)
```

The classification results can be shown both graphically and numerically. Here is a prediction plot for the results.

```{r, fig.width = 9, fig.height = 8}
par(mfrow = c(2, 1))
plotPredictions(res, show.legend = F)
plotPredictions(res, ncomp = 2, show.legend = F)
```

So we can see (the legend has been hidden in order to see all points clearly) that for the model with three components we have three false positives (specificity = 47/50 = 0.94) and three false negatives (sensitivity = 22/25 = 0.88). You can also show the predictions as a matrix with -1 and +1 using method `showPredictions()` or get the array with predicted class values directly as it is shown in the example below (for first 10 rows, different number of components and the first 
classification variable). 

```{r}
show(res$c.pred[31:40, 1:3, 1])
```

You can also get and show the confusion matrix (rows correspond to real classes and columns to the predicted class) for an object with SIMCA results (as well as results obtained with any other classification method, e.g. PLS-DA). In this example the matrix also shows 3 false negatives and 3 false positives we have mentioned before.

```{r}
show(getConfusionMatrix(res))
```

### Class belonging probabilities {-}

In addition to the array with predicted class, the object with SIMCA results also contains an array with class beloning probabilities. The probabilities are calculated depending on how close a particular object is to the the critical limit border. 

To compute the probability we use the theoretical distribution for Q and T^2^ residuals as for computing critical values (defined by the parameter `lim.type`). The distribution is used to calculate a p-value — chance to get object with given residual distance value or larger. The p-value is then compared with signidicance level, $\alpha$, and the probability, $\pi$ is calculated as follows:

$$\pi = 0.5 (p / \alpha) $$

So if p-value is the same as significance level (which happens when object is lying exactly on the acceptance line) the probability is 0.5. If p-value is e.g. 0.04, $\pi = 0.4$, or 40%, and the object will be rejected as a stranger (here we assume that the $\alpha = 0.05$). If the p-value is e.g. 0.06, $\pi = 0.6$, or 60%, and the object will be accepted as a member of the class. If p-value is larger than $2\times\alpha$ the probability is set to 1.

In case of rectangular acceptance area (`lim.type = 'jm'` or `'chisq'`) the probability is computed separately for Q and T^2^ residuals and the smallest of the two is taken. In case of triangular acceptance area (`lim.type = 'ddmoments'` or `'ddrobust'`) the probability is calculated for a combination of the residuals.

Here is how to show the probability values, that correspond to the predictions shown in the previous code chunk.

```{r}
show(res$p.pred[31:40, 1:3, 1])
```

It is also possible to show the probability values as a plot with method `plotProbabilities()`:

```{r, fig.width = 9, fig.height = 8}
par(mfrow = c(2, 1))
plotProbabilities(res, cgroup = c.t$Species)
plotProbabilities(res, ncomp = 2, cgroup = c.t$Species)
```

The plot can be shown for any SIMCA results (including e.g. calibration set or cross-validated results).

### Extreme plot {-}

Extreme plot was [proposed](https://onlinelibrary.wiley.com/doi/abs/10.1002/cem.2506) as a tool for comparing methods for computing of critical limits as well as for optimization of model parameters, e.g. number of components. The idea of the plot is following. For any given $i$ we can compute $\alpha'$ as $\alpha' = i / n$, where $n$ is the number of objects in calibration set, so $i$ is the number of objects outside the acceptance area for the given $\alpha'$. Then we can count number of objects, $m$, in our data that have p-value below the $\alpha'$. 

In ideal case $m = i$, but in fact there will be some small random variation of the $m$ around $i$. This variation can be computed as $±2\sqrt{i (1 - \alpha)}$. For example, if the expected number of actreme objects $i = 5$ and $n = 50$, like in our last model, $\alpha = 5/50 = 0.1$. So we can expect $5±2\sqrt{5 \times 0.9} ≈ 5 ± 4$ objects may have the p-value smaller than $0.1$. 

The `plotExtreme()` method counts the number of objects from calibration set rejected by the model for $i = 1, 2,...,n$ (and corresponding alpha values) and shows the values vs the $i$. Besides that, the statistical limits. If all points are lying within the limits, the classification model works as expected.

Below you will find an example with four extreme plots made for four Versicolor SIMCA models with different methods for estimation of critical limits. 

```{r, fig.width = 9, fig.height = 8}
Ve = iris[51:100, 1:4]

m1 = simca(Ve, 'Versicolor', ncomp = 2, lim.type = 'jm')
m2 = simca(Ve, 'Versicolor', ncomp = 2, lim.type = 'chisq')
m3 = simca(Ve, 'Versicolor', ncomp = 2, lim.type = 'ddmoments')
m4 = simca(Ve, 'Versicolor', ncomp = 2, lim.type = 'ddrobust')


par(mfrow = c(2, 2))
plotExtreme(m1, main = 'JM')
plotExtreme(m2, main = 'Chisq')
plotExtreme(m3, main = 'DD moments')
plotExtreme(m4, main = 'DD robust')
```

## Multiclass classification {-}

Several SIMCA models can be combined to a special object `simcam`, which is used to make a multiclass classification. Besides this, it also allows calculating distance between individual models and a discrimination power — importance of variables to discriminate between any two classes. Let's see how it works.

First we create three single-class SIMCA models with individual settings, such as number of 
optimal components and alpha.

```{r}
m.set = simca(X.set, 'setosa', 3, alpha = 0.01)
m.set = selectCompNum(m.set, 1)

m.vir = simca(X.vir, 'virginica', 3)
m.vir = selectCompNum(m.vir, 2)

m.ver = simca(X.ver, 'versicolor', 3)
m.ver = selectCompNum(m.ver, 1)
```

Then we combine the models into a SIMCAM model object. Summary will show the performance on
calibration set, which is a combination of calibration sets for each of the individual models

```{r]}
m = simcam(list(m.set, m.vir, m.ver))
summary(m)
```

Now we apply the combined model to the test set and look at the predictions.

```{r, fig.width = 9, fig.height = 5}
res = predict(m, X.t, c.t)
plotPredictions(res)
```

In this case the predictions are shown only for the number of components each model found optimal. The names of classes along y-axis are the individual models. Similarly we can show the predicted values.

```{r}
show(res$c.pred[20:30, 1, 1:3])
```

Method `getConfusionMatrix()` is also available in this case.

```{r}
show(getConfusionMatrix(res))
```

There are three additional plots available for multiclass SIMCA model. First of all it is a distance between a selected model and the others.

```{r, fig.width  =9, fig.height = 5}
par(mfrow = c(1, 2))
plotModelDistance(m, 1)
plotModelDistance(m, 2)
```


The second plot is a discrimination power, mentioned in the beginning of the section.

```{r, fig.width = 9, fig.height = 5}
par(mfrow = c(1, 2))
plotDiscriminationPower(m, c(1, 3), show.labels = T)
plotDiscriminationPower(m, c(2, 3), show.labels = T)
```

And, finally, a Cooman's plot showing an orthogonal distance from objects to two selected classes/models.

```{r, fig.width = 9, fig.height = 5}
par(mfrow = c(1, 2))
plotCooman(m, c(1, 3), show.labels = T)
plotCooman(m, c(2, 3), show.labels = T)
```

